---
title: "Bird's Beak Plant Bombus Bee Observation Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data Loading and Preparation

```{r libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(plotly)
library(DT)
library(kableExtra)
library(viridis)
library(gridExtra)
```

```{r data_loading}
df <- read.csv("Collection Observations - Birds Beak-2.csv", stringsAsFactors = FALSE)

# Display the structure of the data
str(df)
head(df)
```

# Data Cleaning and Processing

```{r data_cleaning}
# Clean and process the data
df_clean <- df %>%
  # Remove empty rows
  filter(!is.na(Date) & Date != "") %>%
  
  # Convert date and time columns
  mutate(
    Date = mdy(Date),
    Start.Time = hms(Start.Time),
    End.Time = hms(End.Time),
    
    # Extract numeric values from wind speed
    Wind = as.numeric(Wind),
    
    # Clean temperature (assuming it's in Fahrenheit)
    Temperature = as.numeric(Weather),
    
    # Create a datetime column for start time
    DateTime = ymd_hms(paste(Date, Start.Time)),
    
    # Extract bee counts from Activity columns
    Bombus_on_site = case_when(
      str_detect(tolower(Activity.on.site), "bombus") ~ str_extract_all(Activity.on.site, "\\d+") %>% 
        map_chr(~ifelse(length(.x) > 0, .x[1], "1")),
      str_detect(tolower(Activity.on.site), "none|na") ~ "0",
      TRUE ~ "0"
    ),
    
    Bombus_around = case_when(
      str_detect(tolower(Activity.around), "bombus") ~ str_extract_all(Activity.around, "\\d+") %>% 
        map_chr(~ifelse(length(.x) > 0, .x[1], "1")),
      str_detect(tolower(Activity.around), "none|na") ~ "0",
      TRUE ~ "0"
    ),
    
    # Convert to numeric
    Bombus_on_site = as.numeric(Bombus_on_site),
    Bombus_around = as.numeric(Bombus_around),
    
    # Total bombus count
    Total_Bombus = Bombus_on_site + Bombus_around,
    
    # Create week number
    Week_num = as.numeric(str_extract(Week, "\\d+")),
    
    # Convert video duration to minutes
    Video_Duration_min = case_when(
      str_detect(Video.Duration, "min") ~ as.numeric(str_extract(Video.Duration, "\\d+\\.?\\d*")),
      TRUE ~ 30  # Default to 30 minutes
    )
  ) %>%
  
  # Filter out rows with missing essential data
  filter(!is.na(Date), !is.na(Site..))

# Display cleaned data summary
summary(df_clean)
```

# Exploratory Data Analysis

## Data Overview

```{r data_overview}
# Basic statistics table
overview_stats <- df_clean %>%
  summarise(
    Total_Observations = n(),
    Unique_Sites = n_distinct(Site..),
    Date_Range = paste(min(Date, na.rm = TRUE), "to", max(Date, na.rm = TRUE)),
    Total_Bombus_Observed = sum(Total_Bombus, na.rm = TRUE),
    Avg_Bombus_per_Session = round(mean(Total_Bombus, na.rm = TRUE), 2),
    Sessions_with_Bombus = sum(Total_Bombus > 0, na.rm = TRUE),
    Percent_Sessions_with_Bombus = round((sum(Total_Bombus > 0, na.rm = TRUE) / n()) * 100, 1)
  )

kable(overview_stats, caption = "Dataset Overview") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Bombus Activity Over Time

```{r bombus_time_series}
# Daily bombus activity
daily_activity <- df_clean %>%
  group_by(Date) %>%
  summarise(
    Total_Bombus = sum(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    Avg_Bombus = round(Total_Bombus / Sessions, 2),
    .groups = 'drop'
  )

# Time series plot
p1 <- ggplot(daily_activity, aes(x = Date, y = Total_Bombus)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Daily Bombus Activity",
       x = "Date",
       y = "Total Bombus Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Average bombus per session
p2 <- ggplot(daily_activity, aes(x = Date, y = Avg_Bombus)) +
  geom_line(color = "orange", size = 1) +
  geom_point(color = "darkorange", size = 2) +
  labs(title = "Average Bombus per Session",
       x = "Date",
       y = "Average Bombus Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1, p2, ncol = 1)
```

## Bombus Activity by Site

```{r bombus_by_site}
# Bombus activity by site
site_activity <- df_clean %>%
  group_by(Site..) %>%
  summarise(
    Total_Bombus = sum(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    Avg_Bombus = round(Total_Bombus / Sessions, 2),
    Sessions_with_Bombus = sum(Total_Bombus > 0, na.rm = TRUE),
    Percent_with_Bombus = round((Sessions_with_Bombus / Sessions) * 100, 1),
    .groups = 'drop'
  )

# Bar plot of total bombus by site
p3 <- ggplot(site_activity, aes(x = factor(Site..), y = Total_Bombus, fill = factor(Site..))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Total Bombus Count by Site",
       x = "Site Number",
       y = "Total Bombus Count") +
  theme_minimal() +
  theme(legend.position = "none")

# Percentage of sessions with bombus by site
p4 <- ggplot(site_activity, aes(x = factor(Site..), y = Percent_with_Bombus, fill = factor(Site..))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Percentage of Sessions with Bombus by Site",
       x = "Site Number",
       y = "Percentage of Sessions") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p3, p4, ncol = 1)

# Display site statistics table
kable(site_activity, caption = "Bombus Activity by Site") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Environmental Factors and Bombus Activity

```{r environmental_analysis}
# Temperature vs Bombus activity
temp_analysis <- df_clean %>%
  filter(!is.na(Temperature)) %>%
  group_by(Temperature) %>%
  summarise(
    Avg_Bombus = mean(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    .groups = 'drop'
  )

p5 <- ggplot(df_clean, aes(x = Temperature, y = Total_Bombus)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Temperature vs Bombus Activity",
       x = "Temperature (°F)",
       y = "Total Bombus Count") +
  theme_minimal()

# Wind speed vs Bombus activity
p6 <- ggplot(df_clean, aes(x = Wind, y = Total_Bombus)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Wind Speed vs Bombus Activity",
       x = "Wind Speed",
       y = "Total Bombus Count") +
  theme_minimal()

grid.arrange(p5, p6, ncol = 1)
```

## Time of Day Analysis

```{r time_of_day}
# Convert start time to hours for analysis
df_clean <- df_clean %>%
  mutate(
    Start_Hour = hour(Start.Time),
    Time_Period = case_when(
      Start_Hour < 12 ~ "Morning",
      Start_Hour < 15 ~ "Midday",
      TRUE ~ "Afternoon"
    )
  )

# Bombus activity by time period
time_activity <- df_clean %>%
  group_by(Time_Period) %>%
  summarise(
    Total_Bombus = sum(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    Avg_Bombus = round(Total_Bombus / Sessions, 2),
    .groups = 'drop'
  )

p7 <- ggplot(time_activity, aes(x = Time_Period, y = Avg_Bombus, fill = Time_Period)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Average Bombus Activity by Time Period",
       x = "Time Period",
       y = "Average Bombus Count") +
  theme_minimal() +
  theme(legend.position = "none")

# Hour-by-hour activity
hourly_activity <- df_clean %>%
  group_by(Start_Hour) %>%
  summarise(
    Total_Bombus = sum(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    Avg_Bombus = round(Total_Bombus / Sessions, 2),
    .groups = 'drop'
  )

p8 <- ggplot(hourly_activity, aes(x = Start_Hour, y = Avg_Bombus)) +
  geom_line(color = "purple", size = 1) +
  geom_point(color = "blue", size = 2) +
  labs(title = "Bombus Activity Throughout the Day",
       x = "Hour of Day",
       y = "Average Bombus Count") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(9, 18, 1))

grid.arrange(p7, p8, ncol = 1)
```

## Weather Conditions Analysis

```{r weather_analysis}
# Clean weather conditions
df_clean <- df_clean %>%
  mutate(
    Weather_Clean = case_when(
      str_detect(tolower(Conditions), "sunny") ~ "Sunny",
      str_detect(tolower(Conditions), "overcast") ~ "Overcast",
      str_detect(tolower(Conditions), "cloudy") ~ "Cloudy",
      TRUE ~ "Other"
    )
  )

# Bombus activity by weather condition
weather_activity <- df_clean %>%
  group_by(Weather_Clean) %>%
  summarise(
    Total_Bombus = sum(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    Avg_Bombus = round(Total_Bombus / Sessions, 2),
    Sessions_with_Bombus = sum(Total_Bombus > 0, na.rm = TRUE),
    Percent_with_Bombus = round((Sessions_with_Bombus / Sessions) * 100, 1),
    .groups = 'drop'
  )

p9 <- ggplot(weather_activity, aes(x = Weather_Clean, y = Avg_Bombus, fill = Weather_Clean)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_viridis_d() +
  labs(title = "Average Bombus Activity by Weather Condition",
       x = "Weather Condition",
       y = "Average Bombus Count") +
  theme_minimal() +
  theme(legend.position = "none")

print(p9)

# Display weather statistics table
kable(weather_activity, caption = "Bombus Activity by Weather Condition") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Weekly Progress Analysis

```{r weekly_analysis}
# Weekly bombus activity
weekly_activity <- df_clean %>%
  group_by(Week_num) %>%
  summarise(
    Total_Bombus = sum(Total_Bombus, na.rm = TRUE),
    Sessions = n(),
    Avg_Bombus = round(Total_Bombus / Sessions, 2),
    Sessions_with_Bombus = sum(Total_Bombus > 0, na.rm = TRUE),
    Percent_with_Bombus = round((Sessions_with_Bombus / Sessions) * 100, 1),
    .groups = 'drop'
  )

p10 <- ggplot(weekly_activity, aes(x = Week_num, y = Total_Bombus)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_point(color = "darkgreen", size = 3) +
  labs(title = "Weekly Bombus Activity Trend",
       x = "Week Number",
       y = "Total Bombus Count") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(weekly_activity$Week_num))

print(p10)

# Display weekly statistics table
kable(weekly_activity, caption = "Weekly Bombus Activity Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Data Quality Assessment

```{r data_quality}
# Check for missing data and technical issues
quality_check <- df_clean %>%
  mutate(
    Has_Technical_Issue = str_detect(tolower(Notes), "died|overheated|memory|card|froze"),
    Full_Duration = Video_Duration_min >= 29  # Sessions that recorded close to full 30 minutes
  ) %>%
  summarise(
    Total_Sessions = n(),
    Sessions_with_Issues = sum(Has_Technical_Issue, na.rm = TRUE),
    Percent_with_Issues = round((Sessions_with_Issues / Total_Sessions) * 100, 1),
    Full_Duration_Sessions = sum(Full_Duration, na.rm = TRUE),
    Percent_Full_Duration = round((Full_Duration_Sessions / Total_Sessions) * 100, 1),
    Missing_Bombus_Data = sum(is.na(Total_Bombus)),
    .groups = 'drop'
  )

kable(quality_check, caption = "Data Quality Assessment") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Technical issues by date
technical_issues <- df_clean %>%
  mutate(Has_Technical_Issue = str_detect(tolower(Notes), "died|overheated|memory|card|froze")) %>%
  group_by(Date) %>%
  summarise(
    Total_Sessions = n(),
    Sessions_with_Issues = sum(Has_Technical_Issue, na.rm = TRUE),
    Percent_with_Issues = round((Sessions_with_Issues / Total_Sessions) * 100, 1),
    .groups = 'drop'
  ) %>%
  filter(Sessions_with_Issues > 0)

if(nrow(technical_issues) > 0) {
  kable(technical_issues, caption = "Technical Issues by Date") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

# Interactive Data Exploration

```{r interactive_plots}
# Create interactive plot for bombus activity over time
interactive_plot <- df_clean %>%
  plot_ly(x = ~DateTime, y = ~Total_Bombus, 
          color = ~factor(Site..), 
          size = ~Video_Duration_min,
          text = ~paste("Site:", Site.., "<br>",
                       "Date:", Date, "<br>",
                       "Time:", Start.Time, "<br>",
                       "Bombus Count:", Total_Bombus, "<br>",
                       "Weather:", Weather_Clean, "<br>",
                       "Temperature:", Temperature, "°F", "<br>",
                       "Wind Speed:", Wind),
          hovertemplate = "%{text}<extra></extra>") %>%
  add_markers() %>%
  layout(title = "Interactive Bombus Activity Timeline",
         xaxis = list(title = "Date and Time"),
         yaxis = list(title = "Total Bombus Count"),
         showlegend = TRUE)

interactive_plot
```

# Summary and Recommendations

## Key Findings

```{r summary_findings}
# Calculate key metrics for summary
summary_metrics <- df_clean %>%
  summarise(
    Peak_Activity_Date = Date[which.max(Total_Bombus)],
    Peak_Activity_Count = max(Total_Bombus, na.rm = TRUE),
    Most_Active_Site = Site..[which.max(Total_Bombus)],
    Best_Weather = weather_activity$Weather_Clean[which.max(weather_activity$Avg_Bombus)],
    Optimal_Temperature = round(mean(df_clean$Temperature[df_clean$Total_Bombus > 0], na.rm = TRUE), 1),
    Best_Time_Period = time_activity$Time_Period[which.max(time_activity$Avg_Bombus)],
    .groups = 'drop'
  )

cat("## Key Findings:\n\n")
cat("- **Peak Activity Date:** ", as.character(summary_metrics$Peak_Activity_Date), 
    " (", summary_metrics$Peak_Activity_Count, " bombus observed)\n")
cat("- **Most Active Site:** Site ", summary_metrics$Most_Active_Site, "\n")
cat("- **Best Weather Conditions:** ", summary_metrics$Best_Weather, "\n")
cat("- **Optimal Temperature:** ", summary_metrics$Optimal_Temperature, "°F\n")
cat("- **Best Time Period:** ", summary_metrics$Best_Time_Period, "\n")
cat("- **Overall Success Rate:** ", overview_stats$Percent_Sessions_with_Bombus, "% of sessions had bombus activity\n")
```

## Recommendations for Future Data Collection

Based on the analysis, here are recommendations for your future data collection and AI model training:

1. **Optimal Collection Conditions:**
   - Focus on sunny weather conditions when possible
   - Target the most productive time periods identified in the analysis
   - Consider environmental factors like temperature and wind speed

2. **Site Strategy:**
   - Prioritize sites with higher bombus activity rates
   - Consider expanding monitoring at successful sites

3. **Technical Improvements:**
   - Address camera reliability issues (overheating, memory cards)
   - Ensure consistent 30-minute recording sessions
   - Consider backup equipment for high-activity periods

4. **Data for AI Training:**
   - Current dataset provides good baseline for training data
   - Include environmental metadata with video annotations
   - Consider balancing dataset with negative examples (no bombus activity)

5. **Future Analysis:**
   - Track seasonal patterns as data accumulates
   - Correlate with plant phenology data if available
   - Consider pollen load analysis for bombus specimens

---

*Analysis completed on `r Sys.Date()`*
